### 2016-06-20 paper/algorithms.tex
small paper changes, new plotting scripts

Algorithm~\ref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} could be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.
% 
\subsection{Complexity Analysis}\label{sect:alg_complexity}
%
We now derive the computational complexity of using Algorithm~\ref{alg:refSeries} to solve an inference problem, and compare it to extant strategies. Let the state, adjoint, and parameter variables each have $N$ degrees of freedom, and assume that an `all at once' strategy is used to solve the KKT system. The cost of single linear solve with such a system is $\orderof{3N}^{\gamma}$, where $\gamma$ depends on the linear solver used~\cite{}. The cost of solving for the auxiliary variables, a linear system, is $\orderof{3N}^{\gamma}$, and the cost for computing the supplementary adjoint is $\orderof{6N}^{\gamma}$. \red{Solving the supplementary adjoint can be split into two $N\times N$ systems...}

If one utilizes Algorithm~\ref{alg:refSeries} to solve the inference problem, with $T$ adaptive iterations the total computational cost is,
%
\begin{align}
\label{eq:cost_adapt}
C_{MF} &= \sum_{i=1}^{T}L_{i}(3N)^{\gamma} + 1(3N)^{\gamma} + 2^{\gamma}(3N)^{\gamma} \nonumber \\
&= \sum_{i=1}^{T} (L_i + 1 + 2^{\gamma}) (3N)^{\gamma}
\end{align}
% 
where $L_i$ is the number of nonlinear solver steps needed to solve the KKT system for the low or mixed fidelity model at each adaptive step. The corresponding cost for solving the high fidelity inference problem, with a nonadaptive algorithm, based on a continuation scheme is,
%
\begin{equation}
\label{eq:cost_nadapt}
C_{HF} = \sum_{i=1}^{C}K_{i}(3N)^{\gamma}
\end{equation}
% 
where $K_i$ is the number of nonlinear solver steps needed to solve the KKT system for the high fidelity model at each continuation step, and $C$ is the number of continuation steps.

Comparing Eq.~\eqref{eq:cost_adapt} and~\eqref{eq:cost_nadapt}, we have,
%
\begin{equation}
\label{eq:cost_compare_0}
C_{MF} < C_{HF} \implies \sum_{i=1}^{T} (L_i + 1 + 2^{\gamma}) < \sum\limits_{i=1}^{C}K_{i}
\end{equation}
%
If the low and multi-fidelity models are linear or nonlinear in a very small region $L_i \approx 1$, and we have the following condition,
%
\begin{equation}
\label{eq:cost_compare_1}
T < \frac{\sum\limits_{i=1}^{C}K_{i}}{2 + 2^{\gamma}}
\end{equation}
% 
The bound on $T$ will be strictest if Gaussian elimination is used, with $\gamma = 3$, giving $T < \frac{\sum\limits_{i=1}^{C}K_{i}}{10}$.
%

>>>>

Algorithm~\ref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} could be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.
% 
\subsection{Complexity Analysis}\label{sect:alg_complexity}
%
We now derive the computational complexity of using Algorithm~\ref{alg:refSeries} to solve an inference problem, and compare it to extant strategies. Let the state, adjoint, and parameter variables each have $N$ degrees of freedom, and assume that an `all at once' strategy is used to solve the KKT system. The cost of single linear solve with such a system is $\orderof{3N}^{\gamma}$, where $\gamma$ depends on the linear solver used~\cite{}. The cost of solving for the auxiliary variables, a linear system, is $\orderof{3N}^{\gamma}$. \red{The parts of the supplementary adjoint corresponding to the primary and auxiliary variables can be solved separately, and thus the cost of computing the supplementary adjoint is also $\orderof{3N}^{\gamma}$.}

If one utilizes Algorithm~\ref{alg:refSeries} to solve the inference problem, with $T$ adaptive iterations the total computational cost is,
%
\begin{align}
\label{eq:cost_adapt}
C_{MF} &= \sum_{i=1}^{T}L_{i}(3N)^{\gamma} + 1(3N)^{\gamma} + 2(3N)^{\gamma} \nonumber \\
&= \sum_{i=1}^{T} (L_i + 3) (3N)^{\gamma}
\end{align}
% 
where $L_i$ is the number of nonlinear solver steps needed to solve the KKT system for the low or mixed fidelity model at each adaptive step. The corresponding cost for solving the high fidelity inference problem, with a nonadaptive algorithm, based on a continuation scheme is,
%
\begin{equation}
\label{eq:cost_nadapt}
C_{HF} = \sum_{i=1}^{C}K_{i}(3N)^{\gamma}
\end{equation}
% 
where $K_i$ is the number of nonlinear solver steps needed to solve the KKT system for the high fidelity model at each continuation step, and $C$ is the number of continuation steps.

Comparing Eq.~\eqref{eq:cost_adapt} and~\eqref{eq:cost_nadapt}, we have,
%
\begin{equation}
\label{eq:cost_compare_0}
C_{MF} < C_{HF} \implies \sum_{i=1}^{T} (L_i + 3) < \sum\limits_{i=1}^{C}K_{i}
\end{equation}
%
If the low and multi-fidelity models are linear, or nonlinear in a very small region, then $L_i \approx 1$, and we have the following condition,
%
\begin{equation}
\label{eq:cost_compare_1}
T < \frac{\sum\limits_{i=1}^{C}K_{i}}{4},
\end{equation}
% 
for the adaptive algorithm to be less costly than solving the high-fidelity inverse problem.
%The bound on $T$ will be strictest if Gaussian elimination is used, with $\gamma = 3$, giving $T < \frac{\sum\limits_{i=1}^{C}K_{i}}{10}$.
%

### 2016-06-20 paper/numerical_experiments.tex
small paper changes, new plotting scripts

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. For convenience of notation in this section, we drop subscripts so that $\Psi=\Psi_{LF}$, $Q=Q_{HF}$, and $U=U_{HF}$; for our numerical experiments, we have $Q_{LF}\subset Q$ and $U_{LF}=U$. We note that the error estimate in Equation \ref{eq:finErrExp} can be equivalently written as

@@

The unknown parameters we wish to infer correspond to the forcing field, so that $f(q)=q$. For the low-fidelity model, the inverse problem is linear (the inferred parameters are linear in the observations). Observations consisting of the state at three points in the domain are artificially generated by running the high-fidelity model on a finer mesh with the true forcing field
%
\begin{equation}
f_{true}(x_1,x_2)=
\begin{cases}
1.0 & \textrm{if }(x_1,x_2)\in[0.125,0.375]\times[0.125,0.375] \\
0.8 & \textrm{if }(x_1,x_2)\in[2.375,2.625]\times[0.375,0.625] \\
0 & \textrm{otherwise}.
\end{cases}
\end{equation}
%
The locations of the observations and the region $\Omega_I$ over which the QoI is calculated are shown in Figure~\ref{fig:baseSetup}. Since the inverse problem is ill-posed, we use Tikhonov regularization~\cite{EngHanNeu00}; the regularization term is $\frac{\beta}{2}\int_\Omega \|\nabla f(q)\|_2^2\:\textrm{d}A$, where $\beta=10^{-5}$ is the regularization coefficient. 
%
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/setup_3_3.pdf}
\caption{Locations of the observations and the QoI region. \red{Aaaah spatial coordinates are not $x$ and $y$! ($y$ is one of the auxiliary variables...)}}
\label{fig:baseSetup}
\end{figure}
%

For the numerical simulations, we use the finite element method (FEM), employing a continuous Galerkin formulation with Lagrange elements. We use the \texttt{libMesh} library~\cite{libMeshPaper} for the FEM calculations. \red{The library offers easy calculation of adjoint systems, error estimates and subdomain restricted variables.}\footnote{But we don't use Libmesh's error estimates or restrict variables to subdomains...} The domain is discretized by a regular mesh of quadrilaterals, with 50 and 250 elements along the short and long boundaries, respectively, for a total of 12,500 elements, resulting in a total of 38,403 degrees of freedom. The diffusion coefficient is chosen so that the cell P\'{e}clet number never exceeds 0.1, and thus no stabilization is required.
%
%------------------------------------------------------------%
\subsubsection{Adaptive Model Refinement Results} \label{sec:cdvcdrBaseRef} 
%------------------------------------------------------------%
%
We now present the results for solving the inference problem using Algorithm~\ref{alg:refSeries}. Once the QoI error estimate is calculated using Equation~(\ref{eq:finErrExp}), the error estimate is then decomposed into local contributions, as described in Equation~(\ref{eq:basisblame}). At each iteration, based on this decomposition, we choose the basis functions with the largest error contributions until an additional 5\% of the elements has been marked for refinement. This is repeated until the estimated absolute relative error in the QoI, is less than $1\%$.

@@

Figure~\ref{fig:baseErr} shows the true and estimated absolute relative errors in the QoI for the various mixed-fidelity models generated by Algorithm~\ref{alg:refSeries}. In this case, we see that the QoI error of only $1\%$ is attained with a mixed-fidelity model where the high-fidelity model is used in only about $10\%$ of the domain. We note that there is no guarantee that either the error in the QoI or the relative error in the error estimate will decrease monotonically as more of the domain is refined.
%
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{baseSeries/err_est.pdf}
\caption{True and estimated absolute relative error in QoI, plotted as a function of the percentage area of the domain in which the high-fidelity convection-diffusion-reaction model is used.\red{does this even have enough points to warrant a graph? or would a table be better?}}
\label{fig:baseErr}
\end{figure}
%

@@

The error decomposition for three increasingly large, nested QoI regions $\Omega_I$ given the same set of observations is shown in Figure \ref{fig:qoiStudy}. The bottom row gives the baseline case presented in Section~\ref{sec:cdvcdrBaseRef}, although here the basis functions with the largest $5\%$ of the error are choesn, so the proportion of additional refined elements in each iteration is slightly larger. Although refinement is still most important around the data point closest to $x_1=0$, as the QoI region expands the other two data points become more important in that the error decomposition suggests refinement around them earlier. As the QoI region expands, it is also more clearly noticeable that refinement is not equally important in all parts of the QoI region.

@@

%
\subsection{Constant vs Field Parameters}
%
\subsubsection{Problem Setup}
%
\subsubsection{Results}
%
%bigger steps in 3D problem because qoi region bigger?
\subsection{Convection-Diffusion(-Highly Nonlinear Reaction)}
%
\subsubsection{Results}
%
We consider the convection-diffusion-reaction term described in Section \ref{sec:cdvcdr}, with a reaction term $k_r=-616$ in the high-fidelity model. This reaction term is large enough that the Newton solver will not converge with a zero initial guess. We first solve the inverse problem with the low-fidelity model ($k_r=0$), and then use a simple continuation approach, using the solution at one value of $k_r$ as the initial guess for the next.\footnote{not arclength continuation, which is more difficult to implement, though libMesh appears to have a class for such continuation, which we can use if this seems like a viable direction} We increase the reaction term in increments of $\Delta k_r=100$, and halve the increment each time the step is too large (Newton solver does not converge at the next $k_r$ value). From $k_r=0$ to $k_r=-616$, this results in 9 continuation steps being taken.
%
\subsubsection{Computational Complexity}
%
We compare the complexity of Algorithm~\ref{alg:refSeries} and the continuation method for the high fidelity problem, in the context of the analysis developed in section~\ref{sect:alg_complexity}. For the high fidelity problem, with  with 9 continuation steps, we have $\sum\limits_{i=1}^{C} K_i=30$.

@@

In other words, on using Algorithm~\ref{alg:refSeries} for a highly nonlinear problem, even with the worst case linear solver, one can get to $<1\%$ error in the QoI, with a $20\%$ reduction in computational cost, while avoiding any need for continuation to handle nonlinearities.

>>>>

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. For convenience of notation in this section, we drop subscripts so that $\Psi=\Psi_{LF}$, $Q=Q_{HF}$, and $U=U_{HF}$; for our numerical experiments, we have $Q_{LF}\subseteq Q$ and $U_{LF}=U$. Note that the error estimate in Equation \ref{eq:finErrExp} can be equivalently written as

@@

The unknown parameters we wish to infer correspond to the forcing field, so that $f(q)=q$. For the low-fidelity model, the inverse problem is linear (the inferred parameters are linear in the observations). Observations consisting of the state at three points in the domain are artificially generated by running the high-fidelity model on a finer mesh with the true forcing field
%
\begin{equation}
f_{true}(x_1,x_2)=
\begin{cases}
1.0 & \textrm{if }(x_1,x_2)\in[0.125,0.375]\times[0.125,0.375] \\
0.8 & \textrm{if }(x_1,x_2)\in[2.375,2.625]\times[0.375,0.625] \\
0 & \textrm{otherwise}.
\end{cases}
\end{equation}
%
The locations of the observations and the region $\Omega_I$ over which the QoI is calculated are shown in Figure~\ref{fig:baseSetup}. Since the inverse problem is ill-posed, we use Tikhonov regularization~\cite{EngHanNeu00}; the regularization term is $\frac{\beta}{2}\int_\Omega \|\nabla f(q)\|_2^2\:\textrm{d}A$, where $\beta=10^{-5}$ is the regularization coefficient. 
%
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/setup_3_3.pdf}
\caption{Locations of the observations and the QoI region.}
\label{fig:baseSetup}
\end{figure}
%

For the numerical simulations, we use the finite element method (FEM), employing a continuous Galerkin formulation with Lagrange elements. We use the \texttt{libMesh} library~\cite{libMeshPaper} for the FEM calculations. \red{The library offers easy calculation of adjoint systems, error estimates and subdomain restricted variables.}\footnote{But we don't use Libmesh's error estimates or restrict variables to subdomains...} The domain is discretized by a regular mesh of quadrilaterals, with 250 and 50 elements along the $x_1$ and $x_2$ directions, respectively, for a total of 12,500 elements, resulting in 12,801 degrees of freedom per variable. The diffusion coefficient is chosen so that the cell P\'{e}clet number never exceeds 0.1, and thus no stabilization is required.
%
%------------------------------------------------------------%
\subsubsection{Adaptive Model Refinement Results} \label{sec:cdvcdrBaseRef} 
%------------------------------------------------------------%
%
We now present the results for solving the inference problem using Algorithm~\ref{alg:refSeries}. Once the QoI error estimate is calculated using Equation~(\ref{eq:finErrExp}), the error estimate is then decomposed into local contributions, as described in Equation~(\ref{eq:basisblame}). At each iteration, based on this decomposition, we choose the basis functions with the largest error contributions until an additional 5\% of the elements has been marked for refinement. This is repeated until the estimated absolute relative error in the QoI, is less than $1\%$.

@@

Figure~\ref{fig:baseErr} shows the true and estimated absolute relative errors in the QoI for the various mixed-fidelity models generated by Algorithm~\ref{alg:refSeries}. In this case, we see that the QoI error of only $1\%$ is attained with a mixed-fidelity model where the high-fidelity model is used in only about $10\%$ of the domain. We note that, in general, there is no guarantee that either the error in the QoI or the relative error in the error estimate will decrease monotonically as more of the domain is refined.
%
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/err_est.pdf}
\caption{True and estimated absolute relative error in QoI, plotted as a function of the percentage area of the domain in which the high-fidelity convection-diffusion-reaction model is used.}
\label{fig:baseErr}
\end{figure}
%

@@

The error decomposition for three increasingly large, nested QoI regions $\Omega_I$ given the same set of observations is shown in Figure \ref{fig:qoiStudy}. The bottom row gives the baseline case presented in Section~\ref{sec:cdvcdrBaseRef}, although here we choose the basis functions $i$ whose error $\varepsilon_i$ are among the largest $5\%$, so the proportion of additional refined elements in each iteration is slightly larger. Although refinement is still most important around the data point closest to $x_1=0$, as the QoI region expands the other two data points become more important in that the error decomposition suggests refinement around them earlier. As the QoI region expands, it is also more clearly noticeable that refinement is not equally important in all parts of the QoI region.

@@

%
\subsection{Constant vs Field Parameters}
%
It is not necessary for the low- and high-fidelity models to differ in the physics included. In this section, we consider two models which differ in the space to which the parameter belongs, with the low-fidelity model having fewer degrees of freedom.
%
\subsubsection{Problem Setup}
%
We consider the same high-fidelity model as in Section~\ref{sec:cdvcdr}:
\begin{equation}
k_d\nabla^2 u - \vec{V}\cdot\nabla u + k_ru^2= f(q),\quad q\in U,
\end{equation}
with the same diffusion coefficient $k_d = 0.1$  and reaction coefficient $k_r = -4.2$. The low-fidelity model
\begin{equation}
k_d\nabla^2 u - \vec{V}\cdot\nabla u + k_ru^2= f(q),\quad q\in\R
\end{equation}
differs from the high-fidelity model only in that the parameter $q$ is a constant instead of a field. Then the intermediate mixed-fidelity models have parameter fields which are non-constant in only portions of the domain. For ease of implementation, we require that the resulting parameter field remain continuous at the interface between the low-fidelity and high-fidelity subdomains, although this constraint is not necessary for the theory to hold. The domain, mesh, boundary conditions, and velocity field, as well as the observations, unknown parameters to be inferred, and QoI, remain the same as described in Section~\ref{sec:cdvcdr}. As the inverse problem is ill-posed, except for perhaps in the case where the low-fidelity model is used throughout the domain, regularization is added; the Tikhonov regularization term is $\frac{\beta}{2}\int_\Omega \|\nabla f(q)\|_2^2+f(q)^2\:\textrm{d}A$, where $\beta=10^{-3}$ is a regularization coefficient.
%
\subsubsection{Adaptive Model Refinement Results}
%
As with the previous examples in Section~\ref{sec:cdvcdr}, the decomposition of the error estimate (\ref{eq:basisblame}) is used to select additional regions of the domain in which to use the high-fidelity model. The number of degrees of freedom in the inverse problem increases with the proportion of the domain in which the high-fidelity model is used and the unknown parameter allowed to be a field. With each iteration, an additional $10\%$ of the elements are marked for refinement. This is repeated until the estimated absolute relative error in the QoI, is less than $1\%$.

Figure~\ref{fig:baseRef} shows the local error contributions, as well as the subdomains where the low- and high-fidelity models are used, for the series of mixed-fidelity models thus generated. Each linear Langrange basis function's contribution is plotted at its nonzero node.

%
\subsection{Convection-Diffusion(-Highly Nonlinear Reaction)}
%
\red{Not sure should even keep this section...if we split the supplementary adjoint solve, then the condition for adaptivity being more efficient doesn't depend on the linear solver...we don't need this section to make a cost argument if we have the 3D example...having this section seems to be overkill/redundant if all we want to say is that 'sometimes we need continuation for the HF inverse problem'...}
%
\subsubsection{Results}
%
We consider the convection-diffusion-reaction term described in Section \ref{sec:cdvcdr}, with a reaction term $k_r=-616$ in the high-fidelity model. This reaction term is large enough that the Newton solver will not converge with a zero initial guess. We first solve the inverse problem with the low-fidelity model ($k_r=0$), and then use a simple continuation approach, using the solution at one value of $k_r$ as the initial guess for the next.\footnote{not arclength continuation, which is more difficult to implement, though libMesh appears to have a class for such continuation, which we can use if this seems like a viable direction} We increase the reaction term in increments of $\Delta k_r=100$, and halve the increment each time the step is too large (Newton solver does not converge at the next $k_r$ value). From $k_r=0$ to $k_r=-616$, this results in 9 continuation steps being taken.
%
\subsubsection{Computational Complexity}
%
We compare the complexity of Algorithm~\ref{alg:refSeries} and the continuation method for the high fidelity problem, in the context of the analysis developed in section~\ref{sect:alg_complexity}. For the high fidelity problem, with  with 9 continuation steps, we have $\sum\limits_{i=1}^{C} K_i=30$.

@@

In other words, on using Algorithm~\ref{alg:refSeries} for a highly nonlinear problem, even with the worst case linear solver, one can get to $<1\%$ error in the QoI, with a $20\%$ reduction in computational cost, while avoiding any need for continuation to handle nonlinearities. 
%
\subsection{Convection-Diffusion(-Reaction) in 3D} 
%previous 2D examples were toy problems were good illustrate something (error breakdown patterns), but too simple to really see any computational cost differences
%bigger steps in 3D problem because qoi region bigger?
